
\vspace*{16pt}
\chapter{THE DIFFERENT KINDS OF SOUNDS}
\section{Vowels}

Besides the five normal vowels--а, е, и, о, у--р can also rank as a vowel when
it is (1) between to consonants, or (2) at the beginning of a word before a
consonant; in these cases it is strongly rolled as in Scotland, e.g.

1. С\textdoublegrave{р}бин=\emph{a Serb} (\emph{masc.}) ;
т\`{р}говац=\emph{merchant} ; чв\trc{р}ст=\emph{firm}.\foot{Even in words of
foreign origin, e.g. трпез\`{а}рија=\emph{dining-room}, from the Greek
\textgreek{τραπεζάριον}.}

2. \textdoublegrave{р}ђа=\emph{rust} ; \`{р}вати се=\emph{to wrestle} ;
\textdoublegrave{р}ђав=\emph{bad}.

р  very seldom occurs as a vowel-sound before or after a vowel; when it does it
is  indicated by two dots, e.g.
\newpage

г\"{р}\'{о}це (3 syllables) = \emph{throat} (diminutive).

за\`{\"{р}}ђати (4 syllables) = \emph{to become rusty}.

All vowels, including р, may be either short or long.

\smallskip
{\hfill \large \textbf{`Movable A'} \hfill}

Particular mention must be made of what is knows as the `movable a'. In Serbian
only the following four groups of consonants are possible at the end of words :
ст, шт, зд, жд; when a words would end in any other group than these, an a is
inserted in the \emph{nom, sing.}, but disappears in the other cases where the
word naturally ends in a vowel; but in the \emph{gen. plur.} the a reappears in
these words, a phenomenon caused by the fact that the invariable long
\emph{final} a of this case is of comparatively modern origin.  E.g.

\emph{Nom. sing.} к\`{о}нац=\emph{cotton, cotton, thread}

\emph{Gen. sing.} к\'{о}нца.

\emph{Gen. plur.} к\`{о}н\={а}ц\={а}

It is very frequent in the \emph{nom, sing. masc.} of adjectives, e.g.
ж\'{е}дан (\emph{masc})=\emph{thirsty}, but ж\'{е}дна (\textit{fem.}).

In the case of foreign words practice varies ; thus one finds both
ф\textdoublegrave{а}кт and ф\textdoublegrave{а}кат. 

\smallskip
{\hfill \large \textbf{Final л and о} \hfill}

Final л of a syllable, and especially of a word, very frequently becomes о.
In words which originally ended in -ол in the \emph{nom. sing.} the two о's
then combine into one long vowel, but the л reappears in the othercases, e.g.

в\trc{о} (\emph{m.})=\emph{ox}, gen. sing. в\'{о}ла. 

ст\trc{о} (\emph{m.})=\emph{Table}, gen.sing. ст\`{о}ла.

с\trc{о}=\emph{salt}, gen.sing. с\`{о}ли, the nom. sing. of which was originally
вол, стол, and сол.

In other cases the л appears as о after another vowel when
final, reappearing in other cases, e.g.

б\textdoublegrave{е}о=\emph{white} (nom. sing, masc), but б\'{е}ла
=\emph{white} (nom. sing, fem.), б\trc{е}ли=\emph{white} (nom. plur. masc).
This phenomenon occurs most frequently in the past participle of the verbs, e.g.

\`{и}мао=\emph{(he) had} (masc. sing.), but \`{и}мала=\emph{(she) had} (fem. sing.). 

\pagebreak

It may also occur in the middle of words when n. is at the end
of a syllable, e.g.

се\`{о}ба=\emph{migration} (originally селба).

вл\'{а}даоца, gen. sing. of вл\`{а}далац=\emph{ruler} (e,g, \emph{king}).

Cf. also Бе\`{о}град=\emph{Belgrade} (lit. \emph{the white city}, originally Бел-град).

\section{Consonants}

The consonants, according to the manner of their
articulation, fall into the two groups :

1. Voiced : б, в, г, д, ђ, ж, з, џ.

2. Voiceless : п, ф, к, т, ћ, ш, с, ч, ц, х.

\smallskip
{\hfill \textit{Rule of the assimilation of Consonants} \hfill}

When a voiced and a voiceless consonant come together,\linebreak
assimilation takes place, i.e. both must be either voiced \linebreak 
or voiceless : (1) a voiceless consonant becomes voiced before
a voiced consonant, and (2) vice versa, e.g.

\begin{itemize}
    \item[(1)] св\textdoublegrave{а}дба (\emph{f.})=\emph{wedding} is derived
        from сват+ба (св\textdoublegrave{а}т (\emph{m}.)\\
        \hspace*{3in}=\emph{wedding guest})\\
        \`{о}таџбина ({\it f.}) = \textit{fatherland}\quad ,,\qquad,, \quad
        отач+бина (\`{о}тац ({\it m.})\\
        \hspace*{3in}=\emph{father})  
    \item[(2)] с\textdoublegrave{р}пски ({\it adj.}) =
        \emph{Serbian}\qquad,,\qquad,, \quad 
        срб+ски (с\textdoublegrave{р}бин ({\it m.}))\\
        \hspace*{3in}=\emph{Serbian} ({\it m.})\\
        вр\'{а}пци ({\it nom. pl.}) =\emph{sparrows}\qquad,, \quad
        враб + ци\quad(вр\'{а}бац\\
        \hspace*{3in}({\it m.})= \emph{sparrow})
\end{itemize}

Exceptions: д remains before с and ш, e.g.

\qquad\qquad пр\'{е}дседник ({\it m.}) = \emph{president.}

\qquad\qquad одшкринути = \emph{to open slightly.}

в never changes into ф and does not change preceding voiceless consonants, e.g.

\qquad\qquad к\`{о}л\={е}вка ({\it f.}) (not колефка) = \textit{cradle}.

\qquad\qquad кл\'{е}тва ({\it f.}) (not кледва) = \textit{curse}.

\grammarnote{Most Important Phonetic Rules}

I. The gutterals к, г, х are `softened' when followed (1) by е and (2) by и, as follows :
\newpage

1. ({\it a.}) Before е, к changes into ч, г into ж, х into ш, in voc. sing. of masculine nouns, e.g.

Nom. sing., v\trc{у}к ({\it m.}), \emph{wolf}, voc. sing. в\trc{у}че.

\quad,, \quad ,, \qquad б\textroundcap{о}г \masc, \emph{god},\qquad,,\quad,,\quad б\tdg{о}же.

\quad,, \quad ,, \qquad д\tdg{у}х \masc, \emph{spirit},\quad,,\quad,,\quad д\tdg{у}ше.

({\it b}) In the 2nd and 3rd person singular of the aorist tense (cd. p. 187f.) e.g. р\`{е}ћи = \emph{to tell}, т\tdg{р}гнути = \emph{to pull}.

1st p. р\tdg{е}кох, \emph{I told}, \quad 2nd and 3rd p. р\tdg{е}че. 

\quad,,\quad т\tdg{р}гох, \emph{I pulled}, \quad,, \quad \qquad,,\qquad т\tdg{р}же.

({\it c}) In certain words derived from those ending in these consonants, e.g.

др\textroundcap{у}г \masc, \emph{companion} ; др\'{у}жити се, \emph{to keep company}.

к\`{о}нак \masc, \emph{a hostel} ; к\`{о}начити, \emph{to spend the night}.

с\textroundcap{у}х ({\it adj}), \emph{dry} ; с\'{у}шити, \emph{to dry} (transitive). 

2. Before и, г changes into з, к into ц, х into с, in the nom. dat. voc. inst.
loc. pl. of most nouns whose stems end in these consonants, e.g.

б\'{у}брег \masc, \emph{kidney}, nom. voc. pl. б\'{у}брези, dat. inst. loc. б\'{у}брезима. 

в\textroundcap{у}к \masc, \emph{wolf}, nom. voc. pl. в\textroundcap{у}ци, dat. inst. loc. в\textroundcap{у}цима.  

сир\'{о}мах \masc, \emph{poor man}, nom. voc. pl. сир\'{о}маси, dat. inst. loc. сир\'{о}масима.

II. If ц and з are followed by е or и, they become ч and ж, e.g.

з\textroundcap{е}ц \masc, \emph{hare} ({\it masc}), voc. sing. з\textroundcap{е}че, з\`{е}чица \femn, \emph{hare} ({\it fem.}).

кн\trc{е}з \masc, \emph{price}, \qquad,,\quad,,\quad кн\trc{е}же.

\`{о}тац \masc,\emph{father},\qquad,,\quad,,\quad \`{о}че.

III. In the case of verbs whose roots end in г, к, and х, these consonants
coalesce with т of the infinitive ending -ти and form ћ, cf. p. 102.


IV. The palatal consonant j, in such syllables as -ја-, -је-, -ји-, -ју-,
affects most of the non-palatal consonants if they immediately precede it. Such
consonants coalesce with j into one sound, as follows :\\ 
д+ј=ђ, e.g. мл\tdg{а}ђ\={и} ({\it adj.}) = \emph{younger}, derived from млад +
ји\footnote{-ји s the sign of the comparative.}\\ 
т+ј=ћ, e.g. љ\tdg{у}ћ\={и} ({\it adj.}) = \emph{more angry}\quad,,\qquad,,\quad љут\quad\hspace*{-2pt}+ ји

\newpage

\noindent з+ј=ж e.g. б\tdg{р}ж\={и} ({\it adj})= \emph{quicker}, derived from брз\quad+ ји\\
с+ј=ш, e.g. к\tdg{и}ша \femn = \emph{rain}\qquad\qquad,,\qquad ,,\quad кис\quad+ ја\\
ц+ј=ч, e.g. \tdg{у}жичанин \masc = \emph{a native}\\
\hspace*{0.42\linewidth} \emph{of \trc{У}жице},,\qquad ,,\quad ужиц + јанин\\
и+ј=њ, e.g. т\tdg{а}њ\={и} ({\it adj.}) = \emph{thinner} \quad ,,\qquad ,,\quad тан\quad + ји\\
л+ј=љ, e.g. вес\'{е}ље ({\it n.}) = \emph{joy} \quad\qquad,,\qquad ,,\quad весел + је\\
г+ј=ж, e.g. др\tdg{а}ж\={и} ({\it adj.}) = \emph{dearer} \quad,,\qquad ,,\quad драг\quad+ ји\\
к+ј=ч, e.g. ј\tdg{а}ч\={и} ({\it adj.}) = \emph{stronger} \quad ,,\qquad ,, \quad јак\quad + ји\\
х+ј=ш, e.g. т\tdg{и}ш\={и} ({\it adj.}) = \emph{quieter} \quad,,\qquad ,,\quad тих\quad + ји\\

Further, if such consonants are in their turn preceded by з or с, these become
respectively ж and ш, e.g.\\

гр\trc{о}зд \masc = \emph{bunch of grapes}, but гр\trc{о}жђе \neut \emph{grapes} (collective noun), from грозђе = грозд-је.

л\trc{и}ст \masc = \emph{leaf, sheet (of paper)}, but л\trc{и}шће \neut = \emph{leaves} (collective noun), from лисће = лист-је.

Б\tdg{о}сна \femn = \emph{Bosnia}, but Б\tdg{о}шн\={а}к \masc = \emph{a Bosnian} \masc, from Босњак = Босн-јак.\\

Further, when the syllables beginning with ј are immediately preceded by the consonants б, п, в, м, the letter л is inserted and coalesces with ј, forming the consonant љ, e.g.
